---
title: "LCS_Full_Presentation_Analysis"
author: "Carlos Mercado"
date: "February 24, 2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 00 - Prep

Packages required to run this markdown. 

```{r}
# Tidyverse is a popular dialect of R, simplying nearly everything 
# you'd want to do in a statistical programming language. 

if( !require(tidyverse) ){ 
  stop("You need to install tidyverse to continue")
   # Code you need 
   # install.packages("tidyverse") # once per computer 
   # library(tidyverse) # once per R session
} 

# visnetwork allows for easy network data visualization
# while igraph performs graph calculations
if( !require(visNetwork) ){ 
  stop("You need to install visNetwork to continue")
   # Code you need 
   # install.packages("visNetwork") # once per computer 
   # library(visNetwork) # once per R session
} 

if( !require(igraph) ){ 
  stop("You need to install igraph to continue")
   # Code you need 
   # install.packages("igraph") # once per computer 
   # library(igraph) # once per R session
} 

```
# 0 - Data Simulation
 
Problem: A country has identified signs of widespread impact to 
 child development, possibly a disease or toxic substance outbreak.

Data will be at the child level and consist of
    - Hospital where case was identified
    - Child ID 
    - Age
    - Severity
    - Last (up to 5) City, States of Travel: City_State_1, ..., City_State_5

## Generating synthetic random data

```{r}
set.seed(4) # reproducible randomization 

 # child id will be 1,000 random combinations of 5 letters and 5 numbers 
 
child_id <- NULL
for(i in 1:1000){ 
  child_id <- c(child_id, 
                paste0(sample(x = LETTERS, size = 5, replace = FALSE),
                       sample(0:9, 5),
                       collapse = "")
                )
}

 # age will be skewed, truncated, normal distributed between 1-12,
  # with an average of 6.

age <- abs ( rpois(n = 1000, lambda =  6) ) + 1
age[age > 12] <- 1

 # Severity will be poisson distributed with lambda = 1 and minimum of 1.
   # values greater than 3, made to be 1.

severity <- rpois(1000, 1) + 1
severity[severity > 3] <- 1
  
# Generate a data frame to hold the data 

outbreak_tbl <- data.frame(
  child_id = child_id, 
  age = age, 
  severity = severity
)

 # City, State will be randomized at the child level to result in between 2-5
 # cities per child. Only 10 cities across 3 states made available for simplicity.
 # a child will get a random number of locations, and then draw from these cities
 # such that the first few are disproportionately common. 

city_states <- c("Los Angeles, California", 
                 "San Diego, California",
                 "San Francisco, California",
                 "San Jose, California",
                 "Houston, Texas",
                 "Dallas, Texas",
                 "New York City, New York",
                 "Washington, District of Columbia")

child_travel_history <- data.frame(
  child_id = child_id,
  city_state_1 = NA,
  city_state_2 = NA,
  city_state_3 = NA,
  city_state_4 = NA,
  city_state_5 = NA
)

# randomly select between 2-5 cities for a child to have visited
 # if > 5, make 2.
num_cities <- rpois(n = 1000, lambda = 2) + 2
num_cities[num_cities > 5] <- 2

# get a random distribution of cities to have visited 
city_number <- rpois(n = 1000, lambda = 3) + 1
city_number[city_number > 10] <- 1
city_probabilites <- table(city_number)/1000

# for each child
for(i in 1:1000){
  
  # prep 5 maximum cities   
  temp_cities <- rep(NA, 5)  
  
  # get both a random number of cities N and N random cities. 
  selected_cities <- sample(city_states, size = num_cities[i],
         replace = FALSE,
         prob = city_probabilites)
  
  # fill in the number of cities and assign it to that child's row 
  temp_cities[1:length(selected_cities)] <- selected_cities
  
  child_travel_history[i, 2:6] <- temp_cities
  
  }

child_outbreak_tbl <- merge(outbreak_tbl, child_travel_history, by = "child_id")

# save separately 
write.csv(child_outbreak_tbl, "child_outbreak_simulation.csv", row.names = FALSE)  

```

# 1 - Describe Simulated Data

Generate a few basic summaries of the data, such as number of cases,
number of affiliated locations of travel, and average number of affiliations
per child.

```{r}


child_outbreak_tbl <- read.csv("child_outbreak_simulation.csv", 
                               row.names = NULL)
# 1000 children 
nrow(child_outbreak_tbl)

# table of severity
table( child_outbreak_tbl$severity )

# most common affiliations 

sort ( table( unlist(child_outbreak_tbl[, 4:8]) ) )
```

# 2 - Analyze Simulated Data

Visualize the data to identify insights to update understanding and description
of the data. Such as differences in severity across age.

```{r}
# plot of severity across age
ggplot(child_outbreak_tbl,
       aes(x = 1, 
           y = age)) + facet_grid(cols = vars(severity)) +
  geom_violin(position = position_dodge(width = 0.4)) + 
  geom_boxplot(width = 0.2, outlier.alpha = 0, 
               position = position_dodge(width = 0.4)) +
  theme_classic() + scale_y_continuous(breaks = 1:12) + 
  theme(strip.background = element_rect(fill = 'black'),
        strip.text = element_text(colour = 'white'),
        plot.title = element_text(hjust = 0.5),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y = element_text(size = rel(1.5)),
        legend.position = "none")
        
```


# 3 - Develop Predictions

Leverage statistics to modeling the underlying phenomenon. Here, I'll apply 
a community detection algorithm to identify if the outbreak is widespread 
or contained in a specific number of clusters. From there, I'll use a
classification tree to model severity based on other variables.

```{r}

child_travel_list <- split(child_travel_history, child_id)

child_affiliation <- { 
  
 # create an empty affiliation matrix   
 temp_matrix <- matrix(data = rep(0, 10000), nrow = 1000)
 rownames(temp_matrix) <- child_travel_history$child_id
 colnames(temp_matrix) <- city_states
 
 # go to each child id and check every possible city_state
 # if a city_state is in that child id's travel history
 # change the intersection matrix value to be 1 instead of 0.
 for(i in 1:length(child_travel_history$child_id)) { 
   for(j in 1:length(city_states)) { 
     
     if( city_states[j] %in% 
         child_travel_list[[ child_travel_history$child_id[i] ]] ){ 
       temp_matrix[i,j] <- 1
     }
     
   }
 }
 
 temp_matrix
 
} 

# Transpose this matrix to get a 10 x 10 matrix detailing how many 
# children in common two places have. 

place_to_place_matrix <- t(child_affiliation) %*% child_affiliation

# diagonal identifies number of children affilated with each place
# I'll make it zero to simplify the graph.

diag(place_to_place_matrix) <- 0

# Converting my place x place matrix to an igraph
# using igraph's graph.adjacency
place_to_place_graph <- graph.adjacency(place_to_place_matrix,
                                        mode = "undirected")

# Use Girvan-Newman Edge Betweeness to identify natural communities 

edge.betweenness.community(place_to_place_graph)

# Only a single community detected 
  # My randomization was too random; didn't create discernable communities!

# Generate visnetwork version of graph
place_to_place_visnetwork <- toVisNetworkData(place_to_place_graph)




```

Develop a classification tree model that identifies severity as a function
of age and amount of recent travel.

```{r}

outbreak_model_data <- { 
  
  # get a long form of the wide data 
  # while ignoring NAs introduced from children with < 5 cities
  temp_outbreak <- na.omit ( pivot_longer(child_outbreak_tbl, 
                                cols = starts_with("city_state"),
                                values_to = "associated_city")
  ) 
  
  # get the state to be its own column
  temp_outbreak$associated_state <- { 
  # split on ", "
  city_state_list <- strsplit(temp_outbreak$associated_city, 
                              split = ", ") 
  # grab the 2nd item (the state)
  unlist ( 
    lapply(city_state_list, 
           FUN = function(x){ x[[2]] })
  )
  }

  # drop child id, get age, severity, associated_city, associated_state
  
  temp_outbreak[ , c("severity","age","associated_state")]
}

# show output as visnetwork
ctree_model <- rpart(severity ~ . , data = outbreak_model_data)
visTree(object = ctree_model, 
        main = "Outbreak Classification Tree"
)

```
